Verifying Spatial Properties of Array Computations

### 2017-04-17 oopsla17/evaluation.tex
next eval paragraph

\documentclass[10pt,preprint,numbers]{sigplanconf}

\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{color}
\usepackage{graphics}
\usepackage{fancyvrb}
\usepackage{subfigure}
\usepackage{multirow}
\usepackage{minted}
\usepackage{breqn}
\usepackage{enumitem}
%\usepackage{subcaption}
\usepackage{siunitx} % For pretty-printing numeric values and SI units
                     % of measure. e.g., the tabular column type S is
                     % used to print nice-looking tables of numbers.
\sisetup{ % defaults
  group-separator={,},
  group-minimum-digits={3},
  output-decimal-marker={.},
  table-format = 6
}

\usepackage{tikz}
\usepackage{xypic}
\usepackage{natbib}

\usetikzlibrary{calc}

%% http://tex.stackexchange.com/questions/55068/is-there-a-tikz-equivalent-to-the-pstricks-ncbar-command
\tikzset{
    ncbar angle/.initial=90,
    ncbar/.style={
        to path=(\tikztostart)
        -- ($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)
        -- ($(\tikztotarget)!($(\tikztostart)!#1!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztotarget)$)!\pgfkeysvalueof{/tikz/ncbar angle}:(\tikztostart)$)
        -- (\tikztotarget)
    },
    ncbar/.default=0.5cm,
}

\tikzset{round left paren/.style={ncbar=0.4cm,out=110,in=-110}}
\tikzset{round right paren/.style={ncbar=0.4cm,out=70,in=-70}}

\fvset{
  linenos=true,
  fontsize=\footnotesize,
  breaklines=true,
 breakafter=w),
  xleftmargin=\parindent,
  numbersep=1em
}

\usemintedstyle{vs}

\definecolor{darkblue}{rgb}{0.0,0.0,0.5}
\definecolor{darkgreen}{rgb}{0.0,0.4,0.0}
\definecolor{darkdarkgreen}{rgb}{0.0,0.35,0.0}

\usepackage{hyperref}
\hypersetup{
    unicode=false,          % non-Latin characters in Acrobat's bookmarks
    pdftoolbar=true,        % show Acrobat toolbar?
    pdfmenubar=true,        % show Acrobat menu?
    pdffitwindow=false,      % page fit to window when opened
    pdftitle={},    % title
    pdfauthor={}
    pdfsubject={},   % subject of the document
    pdfnewwindow=true,      % links in new window
    pdfkeywords={keywords}, % list of keywords
    colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=darkblue,          % color of internal links
    citecolor=darkblue,        % color of links to bibliography
    filecolor=green,      % color of file links
    urlcolor=blue,          % color of external links
}

\usepackage{cleveref}
\crefname{section}{\S}{\S\S}
\Crefname{section}{\S}{\S\S}
\crefname{subsection}{\S}{\S\S\S}
\Crefname{subsection}{\S}{\S\S\S}
\crefname{figure}{fig.}{}
\Crefname{figure}{Fig.}{}
\crefname{definition}{def.}{}
\Crefname{definition}{Def.}{}

\CustomVerbatimEnvironment{SpecVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}
\CustomVerbatimEnvironment{ExmVerbatim}{Verbatim}{fontsize=\footnotesize,xleftmargin=0.5cm,
xrightmargin=0.2cm,baselinestretch=0.98,numbers=left,numbersep=0.9em,commandchars=\\\{\}}
\CustomVerbatimEnvironment{IVerbatim}{Verbatim}{fontsize=\relsize{-1},xleftmargin=0.5cm,
xrightmargin=0.2cm,commandchars=\\\{\},baselinestretch=0.98,numbersep=0.9em}

\definecolor{darkgreen}{rgb}{0.0,0.5,0.0}
\definecolor{darkpurple}{rgb}{0.6,0.0,0.6}
\definecolor{orange}{rgb}{0.8,0.4,0.0}
\definecolor{darkorange}{rgb}{0.5,0.2,0.0}
\definecolor{marco}{rgb}{0.0,0.3,0.5}
\definecolor{gray}{rgb}{0.2,0.2,0.2}

\newcommand\mdef{\stackrel{\mathclap{\normalfont\mbox{def}}}{=}}

\newcommand{\bn}{\mathbb{N}}

\newcommand{\todo}[1]{\textcolor{blue}{Dom: #1}}
\newcommand{\dnote}[1]{\textcolor{darkpurple}{Dom: #1}}
\newcommand{\mnote}[1]{\textcolor{darkgreen}{Mistral: #1}}
\newcommand{\anote}[1]{\textcolor{red}{Andy: #1}}

\newcounter{block}

\newtheorem{lemma}[block]{Lemma}
\newtheorem{proposition}[block]{Proposition}

\theoremstyle{definition}

\newtheorem{theorem}[block]{Theorem}
\newtheorem{remark}[block]{Remark}
\newtheorem{example}[block]{Example}
\newtheorem{definition}[block]{Definition}

% Writing macros
\newcommand{\ie}{\emph{i.e.}}
\newcommand{\eg}{\emph{e.g.}}

\newcommand{\dimId}{\texttt{dim}}

% Semantics related
\newcommand{\interp}[1]{\llbracket{#1}\rrbracket}

% Syntax macros
\newcommand{\nonterm}[1]{\textit{#1}}
\newcommand{\term}[1]{\texttt{#1}}

\newcommand{\stenRefl}[1]{\term{pointed} \, (\term{dim=}#1)}
\newcommand{\stenFwd}[3]{\term{forward} \, (\term{depth=}#1,
  \term{dim=}#2{#3})}
\newcommand{\stenBwd}[3]{\term{backward} \, (\term{depth=}#1,
  \term{dim=}#2{#3})}
\newcommand{\stenCen}[3]{\term{centered} \, (\term{depth=}#1,
  \term{dim=}#2{#3})}
\newcommand{\irrefl}{\texttt{nonpointed}}

\newcommand{\stenReflS}[1]{\term{pointed} \, (\term{dim=}#1)}
\newcommand{\stenFwdS}[2]{\term{fwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenBwdS}[2]{\term{bwd} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\stenCenS}[2]{\term{cen} \, (\term{depth=}#1,
  \term{dim=}#2)}
\newcommand{\irreflS}{\texttt{np}}

\newcommand{\stenFwdSR}[3]{\term{fwd} (\term{depth=}#1,
  \term{dim=}#2, #3)}
\newcommand{\stenBwdSR}[3]{\term{bwd} (\term{depth=}#1,
  \term{dim=}#2, #3)}
\newcommand{\stenCenSR}[3]{\term{cen} (\term{depth=}#1,
  \term{dim=}#2, #3)}
\newcommand{\stenReflSR}[1]{\term{point} (\term{dim=}#1)}

% SYNTAX OPERATIONS AND PREDICATES
%\newcommand{\neigh}{\textsf{neigh}}
\newcommand{\arrayTy}{\textsf{array}}
\newcommand{\rhsExp}{\textsf{rhsExp}}
\newcommand{\var}{\textsf{var}}

%% VECTOR NOTATIONS
\newcommand{\vect}[1]{\textbf{#1}}

\newcommand{\vtwohs}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{cc}#1 \, & \, #2\end{array}\right]\!}

\newcommand{\vtwoh}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{cc}#1 \; & \; #2\end{array}\right]}

\newcommand{\vthreeh}[3]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{ccc}#1 \; & \; #2 \; & \; #3\end{array}\right]}
\newcommand{\vtwo}[2]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{l}#1\\#2\end{array}\right]}
\newcommand{\vthree}[3]{\setlength{\arraycolsep}{0em}
\left[\begin{array}{l}$#1$\\$#2$\\$#3$\end{array}\right]}
\newcommand{\stwo}[4]
%{\vtwo{#1}{#2}\!\vtwo{#3}{#4}}
{\setlength{\arraycolsep}{0.1em}
\left[\begin{array}{rr}$#1$ & $#3$\\$#2$ & $#4$\end{array}\right]}

\newcommand{\singleEntry}[2]{\textbf{J}_{#2}^{#1}}
\newcommand{\zeroEntry}[2]{\textbf{K}_{#2}^{#1}}

%% OPERATIONS ON SPANS and VECTORS
\newcommand{\containedin}{\sqsubseteq}

%% MODEL
\newcommand{\effdims}[2]{\mathit{constr}(#1)_{#2}}

\newcommand{\trule}[1]{{\footnotesize{(\text{#1})}}}
%\include{results}

%\newcommand{\cons}[3]{\mathbf{C}_{#3}(#1, #2)}
\newcommand{\consName}{\textbf{agree\textprime}}
\newcommand{\cons}[2]{#1 \,\, \consName{} \,\, #2}
\newcommand{\consAName}{\textbf{agree}}
\newcommand{\consA}[2]{#1 \,\, \consAName{} \,\, #2}
\newcommand{\consSub}[3]{#2 \,\, \consAName{}_{#1} \,\,#3}

\title{Verifying Spatial Properties of Stencil Computations using CamFort}
%\titlebanner{DRAFT -- Do not distribute}
\authorinfo{}{}{}
%\preprintfooter{DO NOT DISTRIBUTE}

\begin{document}
\maketitle

\bibliographystyle{plainnat}

\begin{abstract}
  Stencil computations are a common programming pattern in numerical code and
  typically form the core of graphics and scientific computing applications.
  However, they are often complex and unwieldy, involving substantial
  fine-grained manipulation of array indices across multiple arrays and
  dimensions. This requires careful testing, which is almost always done
  manually and then discarded. In this paper, we propose an alternate approach:
  a specification language and automated tool for verifying the spatial
  behaviour of stencils. The specification language is based on the hypothesis
  that stencil computations read from arrays with a regular, fixed pattern
  which can be captured by a simple set of abstract combinators. These
  specifications can be embedded as annotations in source code, against which
  code can be checked for conformance. For legacy code, spatial specifications
  can be inferred, providing documentation and aiding future software
  maintenance.

  We evaluate our specification language and verification tool
  against a corpus of numerical Fortran code ($\sim$ 1 million lines) for which
  we generate $\sim$ 60,000 specifications, showing the vast majority of stencil
  computations indeed have a simple, regular, static shape.
\end{abstract}

%\keywords{program verification, specification, stencils}

\section{Introduction}\label{sec:intro}

\emph{Stencils} are a ubiquitous programming pattern, common in scientific and
numerical computing applications. Informally, a stencil computation computes an
array whose value at each index $i$ is calculated from a \emph{neighbourhood}
of values around $i$ in some input array(s), \eg{}, the Game of Life,
convolutions in image processing, approximations to differential equations. For
example, the following computes the one-dimensional discrete Laplace transform
(an approximation to a derivative) in Fortran:
%
\begin{minted}{fortran}
do i = 1, (n-1)
      b(i) = a(i-1) - 2*a(i) + a(i+1)
end do
\end{minted}
%
Values \texttt{b(i)} are calculated from a neighbourhood of elements around
\texttt{i} in the input array \texttt{a}. In this example, the access pattern
is simple and easily understood. More complex stencil computations are more
prone to errors from simple lexical mistakes. For example, the following is a
small snippet from a Navier-Stokes fluid simulator (based
on~\citet{griebel1997numerical}) in which two arrays are read with different
data access patterns, across two dimensions.
%%
\begin{minted}[firstnumber=20,xleftmargin=2em]{fortran}
du2dx = ((u(i,j)+u(i+1,j))*(u(i,j)+u(i+1,j))+   &
  gamma*abs(u(i,j)+u(i+1,j))*(u(i,j)-u(i+1,j))- &
  (u(i-1,j)+u(i,j))*(u(i-1,j)+u(i,j))-          &
  gamma*abs(u(i-1,j)+u(i,j))*(u(i-1,j)-u(i,j))) &
  /(4.0*delx)

duvdy = ((v(i,j)+v(i+1,j))*(u(i,j)+u(i,j+1))+   &
  gamma*abs(v(i,j)+v(i+1,j))*(u(i,j)-u(i,j+1))- &
  (v(i,j-1)+v(i+1,j-1))*(u(i,j-1)+u(i,j))-      &
  gamma*abs(v(i,j-1)+v(i+1,j-1))*(u(i,j-1)-     &
  u(i,j))) / (4.0*dely)

laplu = (u(i+1,j)-2.0*u(i,j)+u(i-1,j))/delx/delx+ &
  (u(i,j+1)-2.0*u(i,j)+u(i,j-1))/dely/dely

f(i,j) = u(i,j)+del_t*(laplu/Re-du2dx-duvdy)
\end{minted}
%%
The access pattern is much harder to understand than the one-dimensional
Laplace. The miasma of indexing expressions of the form \texttt{var(i$\pm$a,
j$\pm$b)} is not only hard to read, but is prone to simple textual input
mistakes, \eg{}, swapping \texttt{-} and \texttt{+}, missing an indexing term,
or transforming the wrong variable \eg{} \texttt{(i+1,j)} instead of
\texttt{(i,j+1)}.

In practice, the development process for complex stencil computations involves
ad hoc testing to ensure that no transcription mistakes have been made (\eg{},
by visual inspections, or comparison against an manufactured or analytical
solutions~\cite{farrell2010automated}). Typically such testing is discarded
once the code is seen to be correct.

This is not the only information that is often discarded. The ``shape'' of the
indexing pattern is usually the result of choices made in the
numerical-analysis procedure to convert some continuous equations into a
discrete approximation. Rarely are these decisions captured in the source code,
yet the shape of access is usually uniform and is has a perspicuous and concise
description in numerical analysis literature \eg{}, ``\emph{centered in space,
to a depth of 1}'' referring to indexing terms \texttt{a(i)}, \texttt{a(i-1)}
and \texttt{a(i+1)}.

To support the development of correct stencil computations, we propose a
simple, abstract specification language for the data access pattern of
stencils. The shape of the Laplace example stencil is specified in our language
as:
%
\begin{minted}[linenos=false]{fortran}
!=  stencil centered(depth=1, dim=1) :: a
\end{minted}
%
That is, \texttt{a} is accessed with a symmetrical pattern (``centered'') to a
depth of one in each direction in its first dimension. The Navier-Stokes
example has a shape specified as:
%
\begin{minted}[linenos=false,xleftmargin=0em,xrightmargin=0.8em,breakindent=1.9em]{fortran}
!= stencil centered(depth=1,dim=1)*pointed(dim=2)     + centered(depth=1,dim=2)*pointed(dim=1) :: u

!= stencil forward(depth=1,dim=1)                          * backward(depth=1,dim=2) :: v
\end{minted}
%
The specification requires that, over the whole fragment, \texttt{u} is
accessed with a centered pattern to depth of 1 in both dimensions (this is
known as the \emph{five-point stencil}) and \texttt{v} is accessed in a
neighbourhood bounded forwards to depth of $1$ in the first dimension and
backward to a depth of $1$ in the second dimension. The specification is
relatively small and much more abstract compared to the source code.

We provide a verification tool which can check the spatial correctness of
stencil computations against a specification. Furthermore, the tool can
automatically infer specifications and insert these as comments into the source
code. The main features and benefits of our approach are as follows:
%
\begin{itemize}
\item The primary use case is that a programmer writes a specification first
  before the stencil code. Our tool then checks conformance between the two.
  This specify-and-check approach reduces testing effort and future-proofs
  against bugs introduced during refactoring and maintenance.
%
\item A specification concisely captures the access pattern of a stencil and
  turns it into documentation. The inference procedure efficiently produces
  specifications with no programmer intervention, automatically inserting
  specifications at appropriate places in the source code.
%
\item Our specification format is deliberately abstract, with a small number of
  combinators that \emph{do not involve any indexing expressions}, \eg{}
  \texttt{a(i+1,j-1)}. This contrasts with other specification approaches,
  \eg{}, deductive verification tools such as ACSL, where specifications must
  also be in terms of array-indexing expressions. Therefore, any low-level
  mistakes that could be made whilst programming complex indexing code could
  also be made when writing its specification. Our specifications are more
  abstract and lightweight with the aim of aiding adoption by scientists and
  preventing indexing errors.
\end{itemize}

Our verification tool does not target a class of bugs that can be detected
automatically (\emph{push-button verification}). Instead, stencil computation
bugs must be identified relative to a specification of the intended access
pattern.

In this paper, we solely focus on the definition and the usage of the language
and the acoompanying verification tool and omit the underlying mathematical
model and implementation details due to space considerations. We make the
following contributions:
%
\begin{itemize}
\item We introduce a specification language for stencils that captures common
  forms of data access patterns (\cref{sec:lang});
%
\item We provide an implementation of our approach as an extension to
  CamFort~\cite{camfort}, an open-source program analysis tool for Fortran;
%
\item We report on a quantitative study of stencil computations on a corpus of
  numerical Fortran program (\cref{sec:evaluation}), totalling one million
  lines of code. Our tool identifies and infers specifications for roughly
  60,000 stencil computations in the corpus. Approximately 5,000 of the
  stencils we found are non-trivial, corresponding to code which is a possible
  source of errors.
%
\item We give a verification case study of a stencil computation commonly used
  in scientific computing, simulating programming errors (\Cref{sec:jacobi}).
  Our approach detects all possible simulated programming errors for this
  problem.
\end{itemize}
%

\newcommand{\domainVal}{\mathbb{Z}_\infty}
\section{Stencil specifications}\label{sec:lang}

Our specification language is based on the hypothesis that most forms of array
access in numerical code have a fixed, statically-determined access pattern.
For example, the \emph{five-point stencil} on a two-dimensional array reads
from array indices $(i, j)$, $(i-1, j)$, $(i+1, j)$, $(i, j-1)$, and $(i, j+1)$
for all $i, j$ within the inner boundary of the array (to avoid out-of-bounds
access at the edges). We revisit this hypothesis in \cref{sec:evaluation} with
the inference of such regular stencil patterns on our corpus of numerical
programs.

We give various definitions used in the paper then \cref{subsec:syntax}
outlines the syntax of stencil specifications. \cref{sec:eqs} then provides a
syntactic equational theory for specifications which is sound with respect to
the underlying model.

\paragraph{Notation}
%\label{sec:notation}

\renewcommand*{\arraystretch}{0.8}
For the target language, $e$ ranges over expressions (may be impure) and $v$
over its (imperative) variables. %By
%``variable'', we mean in the imperative sense (named binders to memory cells).

\begin{definition}[Induction variable]
  An integer variable is an \emph{induction variable} if it is the control
  variable of a ``\texttt{for}'' loop incremented by $1$ per iteration. A variable
  is interpreted as an induction variable only within the scope of the loop
  body. Throughout, $i, j, k$ range over induction variables.
%TODO: consider mentioning derived, non-based induction variables
% as we did before.
\end{definition}

\begin{definition}[Array subscripts and indices]
  An \emph{array subscript}, denoted $a(\bar{e})$, is an expression
  which reads from an array $a$ at an \emph{index} specified by a
  comma-separated sequence of integer expressions denoted $\bar{e}$ or
  in expanded form as $(e_1, \ldots, e_n)$. An index $e_i$
  is called \emph{relative} if the expression involves an induction
  variable.
  An \emph{absolute index} is a constant integer expression relative to the
  enclosing loop.
\end{definition}

\begin{definition}[Neighbourhood index]
\label{def:neighbour}
  For an array subscript $a(\bar{e})$ an index $e \in \bar{e}$ is a
  \emph{neighbourhood index} if $e$ is of the form $e \equiv i$, $e \equiv i +
  c$, or $e \equiv i - c$, where $c$ is an integer constant. That is, a
  neighbourhood index is a constant translation of an induction variable. (The
  relation $\equiv$ identifies terms up-to commutativity of $+$ and the inverse
  relationship of $+$ and $-$ \eg{}, $(-b) + i \equiv i - b$).
\end{definition}

\texttt{BEGIN: NEEDS REWRITING}
\begin{definition}[Stencil computations]
\label{def:stencil}
  Let $a$ be an array of dimensionality $n$ and $\overline{b}$ a collection of
  arrays of arbitrary (possibly differing) dimensionalities.  A \emph{stencil
  computation} comprises an iteration by a set $I$ of at most $n$ induction
  variables over a subset of the index space of $a$. Each iteration determines
  the elements of $a$ by an assignment $a(\bar{e}) = e_r$ where each $e \in
  \bar{e}$ is either a neighbourhood or constant index.  For all $b \in
  \bar{b}$ and array subscripts $b(\bar{e}')$ that flow to $e_r$, each $e' \in
  \bar{e}'$ is either a neighbourhood or absolute index.

  We refer to the assignment $a(\bar{e}) = e_r$ and all the associated
  statements with dataflow to $e_r$ as the \emph{stencil kernel}.
\end{definition}

We informally describe what it means for a stencil kernel to be consistent with
a specification. This will be algorithmically described in \Cref{sec:analysis}.
Specifications are associated to a stencil kernel and are declared for one or
more array variables $\bar{v}$ within that kernel.  A stencil specification
defines a \emph{region} associated to an array variable $a$ against which a
stencil kernel should be \emph{consistent}. The region defined by a
specification is an $n$-dimensional rectilinear shape in a discrete space
crossing the origin (which represents a collection of indices which are just
induction variables). Consistency is then a two way requirement. All array
subscripts on $a$ contributing to the kernel should be consistent with the
region. Conversely, all parts of the region should have an index corresponding
it.

\texttt{END: NEEDS REWRITING}

\begin{figure}[t]
\vspace{-0.9em}
\begin{align*}
\def\arraystretch{1}
\setlength{\arraycolsep}{0.2em}
\newcommand{\dimTy}{\mathbb{N}_{>0}}
\begin{array}{rl}
\nonterm{specification} ::= & \nonterm{regionDec} \mid \nonterm{specDec} \\
\nonterm{specDec} ::= & \term{stencil} \; \nonterm{spec} \;
                        \texttt{::} \; v \\
\nonterm{regionDec} ::= &  \texttt{region} \; \texttt{::} \; \nonterm{rvar} \; \texttt{=} \;
                         \nonterm{region}\\[0.4em]
%\nonterm{spec} ::= & \nonterm{spatial} \mid \nonterm{temporal}
%\\[1em]
\nonterm{spec} ::= & [\nonterm{mult},] \; [\nonterm{approx},] \; \nonterm{region} \\
\nonterm{mult} ::= & \term{readOnce} \\
\nonterm{approx} ::= & \term{atMost} \; \mid \; \term{atLeast} \\[0.1em]
\nonterm{region} ::= & \nonterm{rvar} \; \\
\multicolumn{2}{l}{\qquad \mid \; \stenRefl{\dimTy}} \\
\multicolumn{2}{l}{\qquad \mid \;\stenFwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
\multicolumn{2}{l}{\qquad \mid \; \stenBwd{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
\multicolumn{2}{l}{\qquad \mid \; \stenCen{\mathbb{N}_{>0}}{\dimTy}{\;[, \texttt{nonpointed}]}} \\
\multicolumn{2}{l}{\qquad \mid \; \nonterm{region} \, \term{+}
  \, \nonterm{region} \; \mid \; \nonterm{region} \; \term{*} \; \nonterm{region}} \\[0.25em]
%\multicolumn{2}{l}{\qquad\qquad \mid \; \nonterm{rvar}}
%\\[0.5em]
%\nonterm{temporal} ::= \; & \term{dependency} \; (v \; \{ , v \}) [, \texttt{mutual}]
%  \\[0.5em]
\nonterm{rvar} ::= \; & [\text{\term{a}-\term{z}$\,$\term{A}-\term{Z}$\,$\term{0}-\term{9}}]+\\[-1em]
\end{array}
\end{align*}
\caption{Specification syntax (EBNF grammar)}
\label{fig:syntax}
\vspace{-0.8em}
\end{figure}

\subsection{Specification syntax}
\label{subsec:syntax}

\Cref{fig:syntax} gives the syntax of stencil specifications, which is detailed
below. The entry point is the \nonterm{specification} production which splits
into either a \emph{region declaration} or a \emph{specification declaration}.
Regions comprise \emph{region constants} which are combined via region
operators \term{+} and \term{*}.

%\paragraph{Region constants}

Region constants specify a finite contiguous region in a single dimension
relative to the origin and are either \term{pointed}, \term{forward},
\term{backward}, or \term{centered}. The region names are inspired by the
scientific terminology, \eg{} the standard explicit method for approximating
PDEs is known as the \emph{Forward Time, Centered Space} (FTCS)
scheme~\citep{dawson1991finite}.

Each region constant has a dimension identifier $d$ given as a positive natural
number.  Each constant except \term{pointed} has a depth parameter $n$ given as
a positive natural number; \term{pointed} regions implicitly have a depth of
$0$.

A \term{forward} region of depth $n$ specifies a contiguous region in dimension
$d$ starting at the origin. This corresponds to specifying neighbourhood
indices in dimension $d$ ranging from $i$ to $i + n$ for some induction
variable $i$. Similarly, a \term{backward} region of depth $n$ corresponds to
contiguous indices from $i$ to $i - n$ and \term{centered} of depth $n$ from $i
- n$ to $i + n$. A \term{pointed} stencil specifies a neighbour index $i$. For
example, the following shows four specifications with four consistent stencil
kernels reading from arrays \term{a}, \term{b}, \term{c} and \term{d}:
%%
\begin{minted}{fortran}
!= stencil backward(depth=2,dim=1) :: a
e(i, 0) = a(i, 0) + a(i-1, 0) + a(i-2, 0)
!= stencil forward(depth=2,dim=1)*pointed(dim=2) :: b
e(i, j) = b(i, j) + b(i+1, j) + b(i+2, j)
!= stencil centered(depth=1,dim=2) :: c
e(i, j) = (c(j-1) + c(j) + c(j+1))/3.0
!= stencil pointed(dim=3) :: d
e(i, j) = d(0, 0, i)
\end{minted}
%%
Every dimension with neighbourhood indices need to be specified. The remaining
non-conforming indices are not included in the specification, \eg{},
specifications on lines $1$ and $7$ leave some dimensions unspecified since they
do not conform to definition neighbourhood indices.

The \term{forward}, \term{backward}, and \term{centered} regions may all have
an additional attribute \term{nonpointed} which marks absence of the origin.
For example, the following is a \term{nonpointed} \term{backward} stencil
%
\begin{minted}{fortran}
!= stencil backward(depth=2, dim=1, nonpointed) :: a
b(i) = a(i-1) + 10*a(i-2)
\end{minted}

\paragraph{Combining regions}

% r+s means that indices must be consistent with either r or s
% r*s means that indices must be consistent with both r and s at the
% same time

% specifications correspond to a set of schemes

% r+s means an index must be generalised by a scheme in either r or s
%       and every scheme in r must have an index which specialises it
%       and every scheme in s must have an index which specialises it

% r*s means an index must be generalised by a scheme in r and a scheme in s
%       and every pairs of schemes in r and s must have an index which
%           specialises both
%

The region operators \term{+} and \term{*} respectively combine regions
permissively and restrictively.

The restrictive combination of two regions, $r \term{*} s$, means that any
indices in the specified code must be consistent with both $r$ and $s$
simultaneously. Dually, for every pair of components in the regions $r$ and $s$
there must be an index consistent with both. For example, the following
\emph{nine-point stencil} has a specification given by the product of two
\texttt{centered} regions in each dimension:
%%
\begin{minted}[breakindent=2.9em]{fortran}
x = a(i, j)   + a(i-1, j)   + a(i+1, j)
y = a(i, j-1) + a(i-1, j-1) + a(i+1, j-1)
z = a(i, j+1) + a(i-1, j+1) + a(i+1, j+1)
!= stencil centered(depth=1, dim=1) * centered(depth=1, dim=2) :: a
b(i, j) = (x + y + z) / 9.0
\end{minted}
%%
This pattern is common in image convolution applications. The specification
ranges over the values that flow to the array subscript on the left-hand side,
and so ranges over the intermediate assignments to \term{x}, \term{y}, and
\term{z}. Each index in the code is consistent with both specifications
simultaneously, \eg{}, \texttt{a(i-1, j+1)} is within the centered region in
dimension $1$ and the centered region in dimension $2$.

The product $r \term{*} s$ can also be thought of as a bounding box over the
two regions $r$ and $s$.

The permissive combination of two regions $r \term{+} s$ means that any indices
in the specified code must be consistent with either of $r$ or $s$. Dually,
every part of region $r$ must have a corresponding index and similarly for $s$
independently. For example, the following gives the specification of a
five-point stencil which is the sum of two compound \texttt{pointed} and
\texttt{centered} regions in each dimension:
%
\begin{minted}{fortran}
!= stencil centered(depth=1, dim=1)*pointed(dim=2)    + centered(depth=1, dim=2)*pointed(dim=1) :: a
b(i,j) = -4*a(i,j) + a(i-1,j) + a(i+1,j) &
                   + a(i,j-1) + a(i,j+1)
\end{minted}
%%
Here the left-hand side of \texttt{+} says that when the second dimension
(induction variable $j$) is fixed at the origin, the first dimension (induction
variable $i$) accesses the immediate vicinity of the origin (to depth of one).
The right hand side of \texttt{+} is similar but the dimensions are reversed.
This reflects the symmetry under rotation of the five-point stencil. As
presented in \cref{sec:evaluation}, this symmetry is common to stencil
computations used in practice, which makes it easier to write specifications.

\paragraph{Region declarations and variables}

Region specifications can be assigned to region variables via
region declarations. For example, the shape of a
``\emph{Roberts cross}'' edge-detection convolution~\cite{davis1975survey}
can be stated:
%%
\begin{minted}{fortran}
!= region :: r1 = forward(depth=1, dim=1)
!= region :: r2 = forward(depth=1, dim=2)
!= region :: robertsCross = r1*r2
!= stencil robertsCross :: a
\end{minted}
This is useful for common stencil patterns, such as the five-point
pattern, as the region can be defined once and reused.
%%
\paragraph{Modifiers}
%%
Region specifications can be modified
by \emph{approximation} and \emph{multiplicity} information
(in \textit{spec} in \Cref{fig:syntax}).

The \texttt{readOnce} modifier enforces that no index appears more
than once (that is, its multiplicity is one). For example, all of
the previous examples could have \texttt{readOnce} added:
%
\begin{minted}{fortran}
!= stencil readOnce, backward(depth=2, dim=1) :: a
b(i+1) = a(i) + a(i-1) + a(i-2)
\end{minted}
%
This specification would be invalid if any of the array subscripts were
repeated. This modifier provides a way to rule out any accidental repetition of
array subscripts.  The notion is similar to that of linear
types~\cite{wadler1990linear}, where a value must be used exactly once. We opt
for the more informative and easily understood name \texttt{readOnce}. This
modifier is optional, so it need not be present even if the stencil is linear.

In some cases, it is useful to give a lower and/or upper bound for a stencil.
This can be done using either the \term{atMost} or \term{atLeast} modifiers.
This is particularly useful in situations where there is a non-contiguous
stencil pattern, which cannot be expressed precisely in our syntax. For example:
%
\begin{minted}{fortran}
!= stencil atLeast, pointed(dim=1)         :: a
!= stencil atMost, forward(depth=4, dim=1) :: a
b(i) = a(i) + a(i+4)
\end{minted}

\section{Equational theory}
\label{sec:eqs}

Specifications have a theory of equality, $\equiv$, providing axiomatic
semantics for the language. This enables users to rewrite their specification
and provides a way to automatically simplify specifications after inference.
Here we only focus on the former and omit mathematical implications of theory.
Additionally, specifications have a theory of approximation, $<:$, that can be
used by the users to weaken specifications for succinctness in the source.

The equality relation $\equiv$ is defined over \emph{region} syntax which gives
the following axioms:
%
\begin{itemize}
\item $\term{+}$ is idempotent, commutative, and associative;
\item $\term{*}$ is commutative, associative, and distributes over
$\term{+}$, \ie{},  $R \; \texttt{*} \; (S \; \texttt{+} \; T) \; \equiv \; (R \;
                       \texttt{*} \; S) \; \texttt{+} \; (R
                       \; \texttt{*} \; T)$.

\item (\emph{overlapping}) Given two region constants $R_1$ and $R_2$
on the same dimension, then some combinations of region constants
can be coalesced into a single region constant $R_3$ if $R_1$ completely
overlaps $R_2$. There are five possible instantiations of the
following rule $R_1 + R_2 \equiv R_3$:
%
\newcommand{\stenGR}[4]{\small{\textit{#4}(\term{depth=}#1,
  \term{dim=}#2, #3)}}
\newcommand{\stenG}[3]{\small{\textit{#3}(\term{depth=}#1,
  \term{dim=}#2)}}
%
\vspace{-0.2em}
\begin{align*}
\hspace{-1em} & \stenGR{n}{d}{p \vee q}{$R_3$} \\[-0.4em]
\hspace{-1em} \equiv \; & \stenGR{n}{d}{p}{$R_1$} \; \texttt{+} \; \stenGR{m}{d}{q}{$R_2$}
\end{align*}
where $n \geq m$ and $p$ and $q$ range
over booleans representing the absence (true) or presence (false)
of \irrefl{} which are combined via disjunction.
We denote the possibilities as $\term{fwd+fwd} \equiv \term{fwd}$, $\term{bwd+bwd} \equiv \term{bwd}$,
$\term{cen+cen} \equiv \term{cen}$, $\term{cen+fwd} \equiv \term{cen}$ and
$\term{cen+bwd} \equiv \term{cen}$. %Commutative of $\term{+}$
%provides symmetric variants.

\item (\textit{overlapping pointed}) \term{pointed} regions coalesce
  with other regions of the same dimension:
%
\vspace{-0.4em}
\begin{align*}
& \stenG{n}{d}{region} \\[-0.4em]
\equiv \; & \stenGR{n}{d}{p}{region} \; \texttt{+} \;
                          \stenReflS{d}
\end{align*}
\vspace{-1.5em}

>>>>

\input{results}

To study the effectiveness of our approach, 
we built a corpus of over 1 million lines of Fortran code from a
range of scientific computing packages: The Unified Model (UM)~\cite{um},
E3ME~\cite{RePEc:aen:journl:2006se-a12}, BLAS~\cite{blas},
Hybrid4~\cite{GBC:GBC635}, GEOS-Chem~\cite{geos-chem}, Navier (based
on \cite{griebel1997numerical}), Computational Physics
ed. 1~\cite{giordano1997computational},
ARPACK-NG~\cite{arpackng}, MUDPACK~\cite{MUD}, Cliffs~\cite{Cliffs}, and
SPECFEM3D~\cite{specfem3d}.

We first examined how frequently stencil computations occur. We parsed
\num{\overalllinesParsed} lines of Fortran code and found that \overalltickAssignPercent\%
(\num{\overalltickAssign}) of statements have a left-hand side as an
array subscript on neighbourhood indices. This supports the idea that
stencil-like computations are common in scientific code.  We then used
the inference procedure of the previous section to generate
specifications for stencils in the corpus to assess the design of the
language.

We would not expect to infer a stencil for each array statement we
found because our analysis restricts the array-subscript-statements
that we classify as a stencil. In fact, we were able to infer a
stencil from \overalltickAssignSuccessPercentOfTickAssign\% of these
statements. A single statement can involve multiple arrays and we
ended up with \num{\overallnumStencilSpecs} specifications. This shows
that we can express a large number of stencil shapes within our
high-level abstraction and validates our initial hypothesis that many
stencil computations have a regular shape.
